Progress Report

### 2023-09-12 Docs/ProgressReport.Rmd
explore local GLMM unidentifiability issue

```{r, message=FALSE}
# overview at estimated latent function
df %>% 
  filter(id %in% rand_id) %>%
  group_by(id, bin) %>%
  summarise(num = sum(Y)) %>%
  left_join(df_est_latent, by = c("id", "bin")) %>%
  # mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
  ggplot()+
  geom_line(aes(x=bin, y=eta_hat, group = id, linetype = "dashed"))+
  geom_point(aes(x=bin, y = num, group = id), size = 0.5)+
  facet_wrap(~id, scales = "free")+
  geom_vline(xintercept = c(465, 545, 555, 1185, 1195, 1205), col = "red",
             linewidth = 0.5, alpha = 0.6)+
  labs(x = "Bin", y = "Estimated latent function/number of active minutes")
```

- Take a closer look at the unidentifiable bins?

I just realized glmer actually used Laplace approximation as well. Could this also be due to numeric stability? In some bins (465, 545, 555, 1185, 1195, 1205), the approximation failed to converge.

Below is the warning messages and one model summary.

For this bin, the problem is with the fixed effect standard error. It is clearly near-zero.

https://www.learn-mlms.com/07-module-7.html

https://rstudio-pubs-static.s3.amazonaws.com/33653_57fc7b8e5d484c909b615d8633c01d51.html

@@

# bins where glmer returned errors
# names(df_bin_lst)[is.na(try_fit_lst)]
```

```{r}
issue_fit <- glmer(Y ~ 1 + (1|id), data = df_bin_lst[['465']],
                             family = binomial)
summary(issue_fit)

# table(df_bin_lst[['5']]$id, df_bin_lst[['5']]$Y)
```

```{r eval=FALSE}
colSums(table(df_bin_lst[['5']]$id, df_bin_lst[['5']]$Y)==0)
colSums(table(df_bin_lst[['465']]$id, df_bin_lst[['465']]$Y)==0)
```

@@

```{r}
mat_est_unique <- matrix(df_est_latent$eta_hat, nrow=N, ncol=n_bin, byrow = F) 
# row index subject, column binned time
fpca_mod <- fpca.face(mat_est_unique, pve = 0.95, argvals = mid, var=T)

@@

@@

- Boundary hitting issues with numeric maximization
- Warning message: Steady solution not reached
- Large variation in both predicted values and scores. The numeric solution from maximum likelihood is obviously not stable. It is very likely to fail to converge.
- Perhaps it can be fixed with numeric options? But I'd rather not go that route because it depend on chance a lot.

```{r}
# Out-of-sample prediction
# maximum observations time: 360, 720, 1080
df_pred <- df_est_latent %>% filter(id %in% rand_id)
df_pred[, 'pred_t360'] <- df_pred[, 'pred_t720'] <- df_pred[, 'pred_t1080'] <- NA
score_out_mat <- array(NA, dim = c(length(rand_id), K, 3))
# dim indexes subject, eigenfunction and max obs time respectively
```

@@

@@

```{r, warning=FALSE}
df_pred %>%
  ggplot()+
  geom_line(aes(x=bin, y=eta_hat, group = id))+
  geom_line(aes(x=bin, y = pred_t360, col = "360", group = id))+
  geom_line(aes(x=bin, y = pred_t720, col = "720", group = id))+
  geom_line(aes(x=bin, y = pred_t1080, col = "1080", group = id))+
  facet_wrap(~id)
```

@@

```{r, warning=FALSE, message=FALSE, results='hide'}
df_pred2 <- df_est_latent %>% filter(id %in% rand_id)
df_pred2[, 'pred_t360'] <- df_pred2[, 'pred_t720'] <- df_pred2[, 'pred_t1080'] <- NA
score_out_mat2 <- array(NA, dim = c(length(rand_id), K, 3))

@@

```{r, warning=FALSE}
df_pred2 %>%
  ggplot()+
  geom_line(aes(x=bin, y=eta_hat, group = id))+
  geom_line(aes(x=bin, y = pred_t360, col = "360", group = id))+
  geom_line(aes(x=bin, y = pred_t720, col = "720", group = id))+
  geom_line(aes(x=bin, y = pred_t1080, col = "1080", group = id))+
  facet_wrap(~id)
```

>>>>

```{r, message=FALSE}
# overview at estimated latent function
df %>% 
  filter(id %in% rand_id) %>%
  group_by(id, bin) %>%
  summarise(num = sum(Y)) %>%
  left_join(df_est_latent, by = c("id", "bin")) %>%
  # mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
  ggplot()+
  geom_line(aes(x=bin, y=eta_hat, group = id))+
  geom_point(aes(x=bin, y = num, group = id), size = 0.5)+
  facet_wrap(~id, scales = "free")+
  geom_vline(xintercept = c(465, 545, 555, 1185, 1195, 1205), col = "red",
             linewidth = 0.5, alpha = 0.6)+
  labs(x = "Bin", y = "Estimated latent function/number of active minutes")
```

#### Take a closer look at the unidentifiable bins

@@

# bins where glmer returned errors
names(df_bin_lst)[is.na(try_fit_lst)]
```

The fitting problem occurs in the following bins: 465, 545, 555, 1185, 1195, 1205. All of them had two warnings: 1) Non-convergence; 2) near-unidenetifiability. Summary results show that standar error estimates of fixed effects (intercept) are very close two zero.

Below is an example of bin 465:

```{r class.source='fold-show'}
issue_fit <- glmer(Y ~ 1 + (1|id), data = df_bin_lst[['465']], family = binomial)

summary(issue_fit)
```

However, this issue seems to disappear after setting nAGO=2. This argument controls the number of nodes in adaptive Gauss-Hermite quadrature approximation. This also seems to be faster and more accurate!

Reference: https://arxiv.org/pdf/2202.07864.pdf

```{r class.source='fold-show'}
issue_fit <- glmer(Y ~ 1 + (1|id), data = df_bin_lst[['465']], family = binomial, nAGQ=2)

summary(issue_fit)
```

<!-- https://www.learn-mlms.com/07-module-7.html -->

<!-- https://rstudio-pubs-static.s3.amazonaws.com/33653_57fc7b8e5d484c909b615d8633c01d51.html -->

So I decided to re-do latent function estimation while adjusting the nAGQ parameter. Below are some findings:

- nAGQ=0 for all bins: no warnings appeared. This is supposed to be fast but not accurate
- nAGQ=1, 2: warnings appeared in some bins.

```{r, eval=FALSE}
# re-fit local GLMM and check numeric issues
try_fit_lst<-list()
for(i in seq_along(df_bin_lst)){
  this_glm <- tryCatch({glmer(Y ~ 1 + (1|id), data = df_bin_lst[[i]], family = binomial, nAGQ=0)},
                       warning=function(w){NA})
  try_fit_lst[[i]] <- this_glm

# bins where glmer returned errors
names(df_bin_lst)[is.na(try_fit_lst)]
lapply(head(try_fit_lst), summary)
```

So perhaps we could try either setting nAGQ = 0 or a variable nAGQ:

Below is the result for nAGQ = 0:

```{r}
# re-define pre_latent funciton with nAGQ = 0
pred_latent_0AGQ <- function(df){
    this_glm <- glmer(Y ~ 1 + (1|id), data = df, family = binomial, nAGQ = 0)
    eta_hat <- predict(this_glm, type = "link")
    df$eta_hat <- eta_hat
    return(df)
}

# re-fit local GLMMs
df_est_latent_0AGQ <- lapply(df_bin_lst, function(x){pred_latent_0AGQ(x)}) 
# head(df_est_latent_0AGQ[[1]])
df_est_latent_0AGQ <- bind_rows(df_est_latent_0AGQ)

# keep midpoint values
df_est_latent_0AGQ <- df_est_latent_0AGQ %>%
  select(-sind, -Y) %>% distinct(.)
```

```{r, message=FALSE}
df %>% 
  filter(id %in% rand_id) %>%
  left_join(df_est_latent_0AGQ, by = c("id", "bin")) %>%
  mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
  ggplot()+
  geom_line(aes(x=sind, y=eta_hat, group = id))+
  geom_point(aes(x=sind, y = Y, group = id), size = 0.5)+
  facet_wrap(~id, scales = "free")+
  # geom_vline(xintercept = c(465, 545, 555, 1185, 1195, 1205), col = "red",
             # alpha = 0.5, linewidth = 0.5)+
  labs(x = "Time", y = "Estimated latent function (probablity scale)")
```

```{r, message=FALSE}
# overview at estimated latent function
df %>% 
  filter(id %in% rand_id) %>%
  group_by(id, bin) %>%
  summarise(num = sum(Y)) %>%
  left_join(df_est_latent_0AGQ, by = c("id", "bin")) %>%
  # mutate(eta_hat = exp(eta_hat)/(1+exp(eta_hat))) %>%
  ggplot()+
  geom_line(aes(x=bin, y=eta_hat, group = id))+
  geom_point(aes(x=bin, y = num, group = id), size = 0.5)+
  facet_wrap(~id, scales = "free")+
  # geom_vline(xintercept = c(465, 545, 555, 1185, 1195, 1205), col = "red",
  #            linewidth = 0.5, alpha = 0.6)+
  labs(x = "Bin", y = "Estimated latent function/number of active minutes")
```

Obviously the lower bound of estimated latent function got a lot higher. Nothing else seems to change much.

For FPCA models below, I'll use the result from nAGQ=0 since it is free of fitting issues.

@@

```{r}
mat_est_unique <- matrix(df_est_latent_0AGQ$eta_hat, nrow=N, ncol=n_bin, byrow = F) 
# row index subject, column binned time
fpca_mod <- fpca.face(mat_est_unique, pve = 0.95, argvals = mid, var=T)

@@

Compared the nAGQ=1:

- The lower bound of mean function went up
- Shapes of PC functions didn't change a lot
- Covariance matrix looks a lot more different! Near-diagonal correlation got higher and off-diagonal got lower.

@@

- Boundary hitting issues with numeric maximization
- Warning message: Steady solution not reached
- Large variation in both predicted values and scores. The numeric solution from maximum likelihood is obviously not stable. It is very likely to fail to converge.
- Perhaps it can be fixed with numeric options? But I'd rather not go that route because it depend on chance a lot. 
- After setting nAGQ=0, the problem got better for some subjects, but not the others.

```{r}
# Out-of-sample prediction
# maximum observations time: 360, 720, 1080
df_pred <- df_est_latent_0AGQ %>% filter(id %in% rand_id)
df_pred[, 'pred_t360'] <- df_pred[, 'pred_t720'] <- df_pred[, 'pred_t1080'] <- NA
score_out_mat <- array(NA, dim = c(length(rand_id), K, 3))
# dim indexes subject, eigenfunction and max obs time respectively
```

@@

# df_pred

@@

```{r, warning=FALSE}
df_pred %>%
  ggplot()+
  geom_line(aes(x=bin, y=eta_hat, group = id))+
  geom_line(aes(x=bin, y = pred_t360, col = "360", group = id))+
  geom_line(aes(x=bin, y = pred_t720, col = "720", group = id))+
  geom_line(aes(x=bin, y = pred_t1080, col = "1080", group = id))+
  facet_wrap(~id,scales = "free")
```

@@

```{r, warning=FALSE, message=FALSE, results='hide'}
df_pred2 <- df_est_latent_0AGQ %>% filter(id %in% rand_id)
df_pred2[, 'pred_t360'] <- df_pred2[, 'pred_t720'] <- df_pred2[, 'pred_t1080'] <- NA
score_out_mat2 <- array(NA, dim = c(length(rand_id), K, 3))

@@

```{r, warning=FALSE}
df_pred2 %>%
  ggplot()+
  geom_line(aes(x=bin, y=eta_hat, group = id))+
  geom_line(aes(x=bin, y = pred_t360, col = "360", group = id))+
  geom_line(aes(x=bin, y = pred_t720, col = "720", group = id))+
  geom_line(aes(x=bin, y = pred_t1080, col = "1080", group = id))+
  facet_wrap(~id,scales = "free")
```
